---
title: "diving"
description: |
  project 6 - diving analysis
author: charlotte friedman
date: December 9, 2025
format: html
execute: 
  echo: false
  warning: false
  message: false
---

### Diving + Shiny!

This project uses Data Science and interactive graphics to explore my performance as a springboard diver. Data comes directly from my DiveMeets profile and includes:

-   Dive codes
-   Dive descriptions
-   High score and average score
-   How many times I’ve competed each dive
-   Dive group (Forward, Back, Reverse, Inward, Twist)

The project uses:

-   R web scraping (rvest)
-   Data cleaning and wrangling (dplyr)
-   Shiny interactive visualization (ggplot + reactive filtering)

### Step 1 - Import and Scrape My Dive Stats

I read data directly from DiveMeets using rvest, which converts HTML table data into an R data frame.

```{r}
#| echo: false
library(tidyverse)
library(rvest)
library(purrr)
library(praise)
library(robotstxt)
library(stringr)

# url

url <- "https://secure.meetcontrol.com/divemeets/system/profile.php?number=67809"

# checking if this is allowed

try(paths_allowed("https://secure.meetcontrol.com/divemeets/system/profile.php?number=67809"))

# 1 - read html and grab all tables on page

page <- read_html(url)

tables <- page |>  html_table(fill = TRUE)

# 2 - see what tables i got

length(tables)

# 3 - view the column names of each table to find the one with my dive stats

map(tables, ~ colnames(.x))


```

### Step 2 - Select and Clean the Dive Table

DiveMeets had some formatting quirks, so I:

-   removed repeated header rows
-   corrected misplaced score columns
-   kept only useful columns

```{r}
#| echo: false

# 4 - select the dives table
dives_raw <- tables[[3]]

# take a quick look !
glimpse(dives_raw)


# renaming cols to something simpler
colnames(dives_raw) <- c("dive_height", "description", "high_score", "avg_score", "n_times", "NAcol")
```

```{r}
#| echo: false

library(tidyverse)
library(stringr)
library(readr)

dives <- dives_raw |> 
  # 1. Drop the header / title rows
  filter(
    dive_height != "Dive Statistics",
    dive_height != "Dive & Height"
  ) |> 
  # 2. Put the numbers in the right places
  #    (on this page: High Score numbers are in avg_score,
  #     Avg Score numbers are in n_times,
  #     # of Times are in NAcol)
  mutate(
    high_score = parse_number(avg_score),   # High Score
    avg_score  = parse_number(n_times),     # Avg Score
    n_times    = parse_integer(NAcol)       # # of Times
  ) |> 
  # 3. Keep only the useful columns
  select(dive_height, description, high_score, avg_score, n_times)

glimpse(dives)

```

### Step 3 - Add Dive Group Categories

Every dive code’s first digit tells the group:\

1 = Forward, 2 = Back, 3 = Reverse, 4 = Inward, 5 = Twist

```{r}
#| echo: false
#adding in dive group column

dives <- dives |> 
  mutate(
    code_first = str_sub(dive_height, 1, 1),
    group = recode(
      code_first,
      "1" = "Forward",
      "2" = "Back",
      "3" = "Reverse",
      "4" = "Inward",
      "5" = "Twist",
      .default = NA_character_
    )
  ) |> 
  filter(!is.na(group))

glimpse(dives)

```

### Extra Credit: Shiny Interactive Plot!

I expanded this project with Shiny, which updates the visualization instantly based on user input.\

Users can choose a dive group and explore my average scores dynamically.

This demonstrates:

-   The Shiny UI + server reactive framework
-   Interactive filtering of *my* real athletic stats
-   ggplot updating automatically based on selections

```{r}

library(shiny)
library(tidyverse)

ui <- fluidPage(
  titlePanel("Interactive Dive Score Explorer"),

  sidebarLayout(
    sidebarPanel(
      selectInput(
        "group_choice",
        "Choose a dive group:",
        choices = sort(unique(dives$group)),
        selected = "Forward"
      )
    ),

    mainPanel(
      plotOutput("scorePlot")
    )
  )
)

server <- function(input, output, session) {

  filtered <- reactive({
    dives |>  filter(group == input$group_choice)
  })

  output$scorePlot <- renderPlot({
    ggplot(filtered(),
           aes(x = reorder(dive_height, avg_score),
               y = avg_score,
               fill = dive_height)) +
      geom_col() +
      coord_flip() +
      labs(
        title = paste("Average Score —", input$group_choice, "Dives"),
        x = "Dive Code",
        y = "Average Score"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
  })
}

shinyApp(ui, server)
```
