---
title: "full text analysis"
description: |
  project 2 - completing a full text analysis
author: charlotte friedman
date: October 8, 2025
format: html
execute: 
  echo: false
  warning: false
  message: false
---

This project analyzes tweets posted by former President Barack Obama using data from the Barack Obama Twitter Archive. The goal is to explore what topics received attention, how messaging changed over time, and how political communication was shaped by hashtags and language patterns. I focus on three linguistic features: references to America, hashtag usage, and mentions of policy areas.

```{r}

library(tidyverse)
library(stringr)

obama_clean <- read_csv("tweets.csv") |>
  mutate(
    # clean and normalize text
    text = str_replace_all(text, "https?://\\S+", ""),
    text = str_to_lower(text),
    text = str_squish(text),
    
    # detect if tweets mention America
    has_america = str_detect(text, "america|american"),
    
    # count hashtags per tweet
    num_hashtags = str_count(text, "#\\w+"),
    
    # extract the first hashtag (if any)
    first_hashtag = str_extract(text, "#\\w+"),
    
    # extract all hashtags
    hashtags = str_extract_all(text, "#\\w+"),
    
    # detect policy-related tweets (simple keyword match)
    policy_related = str_detect(text, "healthcare|education|climate|economy"),

    # convert timestamp to Date
    date = as.Date(timestamp)
  )

```

The frequency of Obama’s tweets changes over time. There are visible spikes during major political moments such as elections, major speeches, and legislative pushes. This suggests Twitter served as a strategic communication tool rather than a constant personal feed, activity increased when public engagement was most important.

```{r}

#graphing!

colnames(obama_clean)

obama_clean |> 
  mutate(date = as.Date(timestamp)) |> 
  count(date) |> 
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  labs(
    title = "Obama Tweets Over Time",
    x = "Date",
    y = "Number of Tweets"
  )
```

Hashtags help reveal Obama’s messaging priorities. The most common hashtags include policy topics such as #obamacare, #education, and #climatechange, emphasizing the administration’s key legislative and advocacy themes. Their high frequency indicates that social media played a role in promoting government initiatives and shaping public conversation around policy values.

```{r}

obama_clean |> 
  unnest(hashtags) |> 
  count(hashtags, sort = TRUE) |> 
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(hashtags, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Hashtags in Obama Tweets",
    x = "Hashtag",
    y = "Count"
  )
```

I used regular expressions to automatically detect whether tweets contained certain political themes. For example, searching for "america" and "american" helps identify nationalistic messaging, while hashtag extraction highlights how Obama’s team used platform-specific tools to coordinate public conversation. The policy-related keyword search allows me to isolate communication focused on healthcare, education, climate issues, and the economy, topics central to Obama’s presidency.

Data Source:\
The data used in this analysis come from the Barack Obama Twitter Archive, collected and published by the U.S. National Archives and Records Administration (NARA).\
Available at: <https://github.com/ObamaWhiteHouse/twitter-archive>
