---
title: "full text analysis"
description: |
  project 2 - completing a full text analysis
author: charlotte friedman
date: October 8, 2025
format: html
execute: 
  warning: false
  message: false
---

```{r}

library(tidyverse)
library(stringr)

obama <- read_csv("tweets.csv")
view(obama)

obama_clean <- obama %>%
  mutate(
    text = str_replace_all(text, "https?://\\S+", ""),   # remove URLs
    text = str_to_lower(text),                           # lowercase
    text = str_squish(text)                              # remove extra spaces
  )

#using required string functions

#detect if tweets mention America
obama_clean <- obama_clean %>%
  mutate(has_america = str_detect(text, "america|american"))

#count hashtags per tweet
obama_clean <- obama_clean %>%
  mutate(num_hashtags = str_count(text, "#\\w+"))

#extract the first hashtag (if any)
obama_clean <- obama_clean %>%
  mutate(first_hashtag = str_extract(text, "#\\w+"))

#using regular expressions

#extract all hashtags
obama_clean <- obama_clean %>%
  mutate(hashtags = str_extract_all(text, "#\\w+"))

#detect tweets mentioning "President" followed by a word (lookbehind)
obama_clean <- obama_clean %>%
  mutate(president_topic = str_extract(text, "(?<=president )\\w+"))

#detect tweets about policies (healthcare, education, etc.)
obama_clean <- obama_clean %>%
  mutate(policy_related = str_detect(text, "healthcare|education|climate|economy"))


```

The line plot showing Obama’s tweets over time reveals clear patterns in his social media activity. Tweet frequency was generally higher during major political periods such as elections and key policy announcements. This suggests that Twitter was used strategically to engage the public during significant events, with noticeable spikes corresponding to campaign seasons and major legislative milestones.

```{r}

#graphing!

colnames(obama_clean)

obama_clean |> 
  mutate(date = as.Date(timestamp)) |> 
  count(date) |> 
  ggplot(aes(x = date, y = n)) +
  geom_line() +
  labs(
    title = "Obama Tweets Over Time",
    x = "Date",
    y = "Number of Tweets"
  )
```

The bar chart of the most frequent hashtags highlights the dominant themes in Obama’s communication strategy. Hashtags such as #obamacare, #education, and #climatechange appear frequently, emphasizing the administration’s focus on healthcare, education, and environmental policy. This pattern demonstrates how social media served as a tool to reinforce political messaging and policy priorities in an accessible, public-facing format.

```{r}

obama_clean |> 
  unnest(hashtags) |> 
  count(hashtags, sort = TRUE) |> 
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(hashtags, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Hashtags in Obama Tweets",
    x = "Hashtag",
    y = "Count"
  )

```
