---
title: "all taylor swift songs"
description: |
  Analysis of all Taylor Swift songs data from TidyTuesday
author: charlotte friedman
date: September 24, 2025
format: html
execute:
  echo: false
  warning: false
  message: false
---

This data set comes from the taylor R package, which compiles a curated collection of Taylor Swift songs including lyrics and audio characteristics. The data originates from Genius and the Spotify API. The visualization displays the trend in average song duration over time, showing the average length of songs each year from before 1960 up through 2020.

```{r}

library(readr) 
taylor_all_songs <- read_csv("taylor_all_songs.csv")
```

```{r}

library(ggplot2)

taylor_all_songs |> 
  ggplot(aes(x = danceability, y = energy)) +
  geom_point(aes(color = album_name), na.rm = TRUE) +
  labs(
    title = "Danceability and Energy of All Taylor Songs",
    x = "Danceability", 
    y = "Energy"
  ) + theme_gray()

```

Data Accessed From:

R for Data Science Online Learning Community (2023). *TidyTuesday: Taylor Swift Songs Dataset* \[Data set\].\
Retrieved from https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-10-17

Original Data Source and Metadata via R Package:

Thompson, W. J. (2023). *taylor: Taylor Swift Songs Data* (GitHub repository).\
Retrieved from https://github.com/jthomasmock/taylor\
This data originates from Genius: https://genius.com/ and Spotify API: https://developer.spotify.com/
