---
title: "visual inference lineup — global temperature trends"
author: "Your Name"
date: "October 30, 2025"
format: html
---

```{r}

#| warning: false
#| message: false
library(tidyverse)

# Working CSV (global temperature anomalies 1880–2023)
global <- read_csv("https://datahub.io/core/global-temp/r/annual.csv")

# Inspect and clean
global <- global %>%
  filter(Source == "GISTEMP") %>%
  rename(Year = Year, temp_anomaly = Mean) %>%
  select(Year, temp_anomaly)
```

## Overview

In this project, I looked at whether there’s a real upward trend in global temperature over time. The data come from the NOAA Global Land–Ocean Temperature Index (GISTEMP).

To test the trend, I used something called a visual inference lineup. Basically, we compare the real data to a bunch of “fake” plots where the data have been shuffled. If people can spot the real plot, that means there’s likely a true trend.

## The Real Data

```{r}

ggplot(global, aes(x = Year, y = temp_anomaly)) +
geom_line(color = "firebrick", linewidth = 1.1) +
geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
labs(
title = "Global Temperature Anomalies (1880–2023)",
subtitle = "Data Source: NOAA / NASA GISTEMP",
x = "Year",
y = "Temperature Anomaly (°C)"
) +
theme_minimal()


```

This line chart shows how each year’s average global temperature compares to a long-term baseline. It looks like the temperatures are going up over time — but we’ll check if that pattern could just happen by chance.

## Simulating the Null Hypothesis

If there’s no real trend, the temperature values shouldn’t be linked to the year. To simulate this, I wrote a function that randomly mixes up the temperature values and makes a new plot.

```{r}
simulate_null_plot <- function(data) {
shuffled <- data %>%
mutate(temp_anomaly = sample(temp_anomaly))

ggplot(shuffled, aes(x = Year, y = temp_anomaly)) +
geom_line(color = "grey50", alpha = 0.8) +
geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dotted") +
theme_minimal() +
labs(x = "Year", y = "Temp Anomaly (°C)")
}
```

## Creating the Lineup

Now I use `map()` to make several of these shuffled plots, then add the real one to compare them all side by side.

```{r}

set.seed(42)

# Create 5 null datasets

null_dfs <- map(1:5, ~global %>%
mutate(temp_anomaly = sample(temp_anomaly), lineup_id = paste0("Null ", .x)))

# Real data labeled separately

real_df <- global %>% mutate(lineup_id = "Real Data")

# Combine all datasets

lineup_df <- bind_rows(null_dfs, real_df)

# Plot all lineups using facet_wrap

ggplot(lineup_df, aes(x = Year, y = temp_anomaly)) +
geom_line(color = "steelblue", alpha = 0.8) +
geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dotted") +
facet_wrap(~lineup_id, ncol = 3) +
theme_minimal() +
labs(
title = "Visual Inference Lineup: Is There a Trend in Global Temperatures?",
subtitle = "Each panel shows a dataset with years vs. temperature anomalies — only one is real.",
x = "Year",
y = "Temperature Anomaly (°C)"
)

```

## What it means

If people can easily pick out the real data from the shuffled ones, that’s strong visual evidence that there’s a true trend. Even without running a formal test, the real plot stands out with a steady rise over time. That makes it clear that global temperatures are really increasing — not just random noise.

## 
References

Buja, A., Cook, D., Hofmann, H., Lawrence, M., Lee, E., Swayne, D. F., & Wickham, H. (2009). *Statistical Inference for Exploratory Data Analysis and Model Diagnostics.* *Philosophical Transactions of the Royal Society A*, 367(1906), 4361–4383.

Wickham, H., Cook, D., Hofmann, H., & Buja, A. (2010). *Graphical Inference for Infovis.* *IEEE Transactions on Visualization and Computer Graphics*, 16(6), 973–979.
